Haskell in Computational Biology

### 2013-05-29 jfp.tex
incorporates AG's comments and Noah's contribution on Senington
this draft might be submittable, although there is one nasty page break

\pagetuningfalse % trying to get page breaks in the best spots

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secrefpair{improving-viterbi}{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds a new technical section (\secref{improving-viterbi})
which describes our experience 
refactoring and improving our implementation of Viterbi's 
\citeyearpar{Viterbi:1967hq}
algorithm.
We~have also simplified the stochastic search as well as other elements of
the code.
Finally, we~have improved and added to explanations, figures, and diagrams.

@@

@@

%
% N.B. very tight paragraph...
%
MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's likelihood is computed
based on frequencies of amino-acid pairs observed in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the rest of
the query sequence, between and around beta strands,
is computed quickly and exactly 
using Viterbi's algorithm.
\ifpagetuning
This likelihood is \emph{conditioned} on the placement.
\else
The exact result is a probability that is
\emph{conditioned} on the placement.
\fi

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing required only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm uses very similar
functions, except for \texttt{nextPt}:
recombination of parent
placements took forty lines of code
and a full day to implement.
{\ifpagetuning\looseness=1 \par\fi}

@@

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version in \mrfy's predecessor,
SMURF.
\ifpagetuning\vadjust{\break}\fi
For example, on a microbenchmark that searches for a
structural motif of 343~nodes in a protein of 2000 amino acids
using only Viterbi's algorithm and no beta-strand information,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made it easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
Errors are estimated from 5~runs.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instances of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.

@@

Research software for
computational biology doesn't have to be fast to be useful.
But~seeing \mrfy\ solve some problems 8~times slower then
SMURF felt like a poor showing for Haskell.
\mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
we~felt sure that with some love and attention, we could make it
more competitive with~C++.
But~before we could try to make \mrfy\ faster, we~had to make it simpler.

@@

To~reduce case analysis even more, we~abstracted over state labels.
\figref{list-viterbi} sketches the structure of the algorithm.
\figref{viterbi-hov4} is the first example in this paper that
shows \emph{all} the cases for~\verb+vee'+.
It~reduces the original 25~cases to just~4.
We~show only a sketch because we didn't use this version in
production.
The difficulty was in memoizing a \verb+vee'+ that takes lists as
arguments. 
We~built and tested specialized memo combinators that work
with list prefixes, but 
we~were never confident that the resulting code would be simple,
correct, and efficient.
Instead, we~used insights from the list-based code to create a new
array-based version:
\begin{itemize}
\item
We~kept the abstraction over \texttt{stateHat}, which we explain
below.
\item
We~got the idea that if we have to compute with integers, we~want
$i$ to \emph{count} the number
of amino acids in the sequence~\texttt{aas}, 
but we want $j$ to \emph{index} into an array of nodes.
\item
From
the \texttt{next} function, we~got the idea that Viterbi's algorithm
computes a minimum edit 
distance.
This idea didn't affect the code directly, but as we experimented with
new variations, it~helped us keep track of the big picture.
\end{itemize}
Armed with these ideas, we~turned back to computing with arrays and
integers.

@@

To~develop a more detailed understanding of performance, we~turned
to GHC's profiler, running Viterbi's algorithm on a single query sequence
of 592~residues, against a
model of 343~nodes.
To~help you understand the results, we~observe that the memo table
for \verb+vee''+ has 610,197 entries: one for each combination of
$\hat s$, $j$, and~$i$.
The~profiler told us that a lot of time was spent in memory
management, and that \mrfy\ allocated 531MB, or 7.85KB
per~entry.
This seems excessive.
%
The profile contained some information we
understood and some we didn't.
\begin{itemize}
\item
The most easily understood source of allocation
is the function \verb+vee''+, which is charged with
allocating about 66MB.
What \verb+vee''+ does is call memo combinators,
so the 66MB might be the memo table.
To~evaluate this hypothesis, we~estimated the amount of memory that
might be consumed by a memo-table entry.

@@

To~try to understand and eliminate the unwanted allocations,
we~explored a many variations on the code, of~which the most ambitious (and
most tedious) was to desugar the list comprehension by~hand and to
replace the standard list operations with operations that participate
in stream fusion \cite{coutts:fusion}.
Nothing~helped.
\end{itemize}
In~desperation, we~tried to diagnose our problems by reading 
Core output from~GHC.
We~were guided by Don Stewart's invaluable blog posts,
and we also used his \texttt{ghc-core} tool, 
but it wasn't enough.
Even though one of us has worked on several optimizing compilers, 
we~just don't have enough experience to use GHC Core to diagnose
allocation problems in a 25-line function.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
to tests, QuickCheck properties, and generators.
Neither Matt nor SMURF includes test code.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
\mrfy's initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
{\ifpagetuning\looseness=1 \par\fi}
\end{itemize}
Like much research software, all three tools were written
 in~haste.
We~have experience modifying the older tools.
ALL THESE TOOLS

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was \emph{too} clever: 
nearly every change resulted in new~segfaults.
{\ifpagetuning\looseness=1 \par\fi}

@@

We~built MRFy quickly, and we expect that
higher-order functions will make MRFy easy to extend.
Each new addition to MRFy's stochastic search has taken at most a day
to implement.
TALK ABOUT RICHARD SENINGTON'S STUFF HERE.

Haskell encourages hasty programmers to slow down.
We~have to get the types right,
which makes it hard to write very large functions.
To~get the code to typecheck, we~have to write type signatures, which
also serve as documentation.
And once the types are accepted by the compiler,
it~is not much more work to write contracts for important functions.
MRFy is still hasty work.
Many types could be simplified;
we're~sure we've missed opportunities for abstraction;
and we know that MRFy's decomposition into modules could be improved.
But despite being hasty programmers, we produced code 
that is easy to understand and easy to extend.
Our hastily written Haskell beats
our hastily written Ruby and~C++.
{\ifpagetuning\looseness=1 \par\fi}

>>>>

\pagetuningtrue % trying to get page breaks in the best spots

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secrefpair{improving-viterbi}{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
would be which
 (\secref{penumbra}).
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds a new technical section (\secref{improving-viterbi})
which describes our experience 
refactoring and improving our implementation of Viterbi's 
\citeyearpar{Viterbi:1967hq}
algorithm.
We~have also simplified the stochastic search as well as other elements of
the code.
Finally, we~have improved and added to explanations, figures, and diagrams.

@@

\ifpagetuning\enlargethispage{1.2\baselineskip}\fi

@@

%
% N.B. very tight paragraph...
%
MRFy treats the beta 
strands in the model
 as ``beads'' which can slide along the query sequence.
A~positioning of the beta strands is called
 a \emph{placement}.
A~placement's likelihood is computed
based on frequencies of amino-acid pairs observed in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of the rest of
the query sequence, between and around beta strands,
is computed quickly and exactly 
using Viterbi's algorithm.
\iffalse
This likelihood is \emph{conditioned} on the placement.
\else
The exact result is a probability that is
\emph{conditioned} on the placement.
\fi

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing required only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm uses very similar
functions, except for \texttt{nextPt}:
recombination of parent
placements took forty lines of code
and a full day to implement.

@@

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version in \mrfy's predecessor,
SMURF.
For example, on a microbenchmark that searches for a
structural motif of 343~nodes in a protein of 2000 amino acids
using only Viterbi's algorithm and no beta-strand information,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
Errors are estimated from 5~runs.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instances of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.

@@

Software for
computational-biology research doesn't have to be fast to be useful.
But~if \mrfy\ solves some problems 8~times slower then
SMURF, that's a poor showing for Haskell.
\mrfy\ spends 98\%\ of its time running Viterbi's algorithm, and
we~felt sure that with some love and attention, we could make it
more competitive with~C++.
But~before we could try to make \mrfy\ faster, we~had to make it simpler.

@@

To~reduce case analysis even more, we~abstracted over state labels.
\figref{list-viterbi} sketches the structure of the algorithm.
\figref{viterbi-hov4} is the first example in this paper that
shows \emph{all} the cases for~\verb+vee'+.
It~reduces the original 25~cases to just~4.
We~show only a sketch because we didn't use this version in
production.
The difficulty was in memoizing a \verb+vee'+ that takes lists as
arguments. 
We~built and tested specialized memo combinators that work
with list prefixes, but 
we~were never confident that the resulting code would be simple,
correct, and efficient.
Instead, we~used insights from the list-based code to create a new
array-based version:
\begin{itemize}
\item
We~kept the abstraction over \texttt{stateHat}, which we explain
below.
\item
We~got the idea that if we have to compute with integers, we~want
$i$ to \emph{count} the number
of amino acids in the sequence~\texttt{aas}, 
but we want $j$ to \emph{index} into an array of nodes.
\item
From
\figref{list-viterbi}'s \texttt{next} function, we~got the idea that
Viterbi's algorithm 
computes a minimum edit 
distance.
This idea didn't affect the code directly, but as we experimented with
new variations, it~helped us keep track of the big picture.
\end{itemize}
Armed with these ideas, we~turned back to computing with arrays and
integers.

@@

To~develop a more detailed understanding of performance, we~turned
to GHC's profiler.
We ran Viterbi's algorithm on a single query sequence
of 592~residues, against a
model of 343~nodes.
The~profiler told us that a lot of time was spent in memory
management, and that \mrfy\ allocated a total of 531MB.
\mrfy's main data structure is the memo table used for \verb+vee''+, 
which has one entry for each combination of
$\hat s$, $j$, and~$i$.
In~this run, that's 610,197 entries.
We~expected \mrfy\ to use an amount of memory proportional to the
size of the memo table.
But the constant of proportionality is 7.85KB, which seems excessive.
We~hoped that the detailed profile would explain, but it
contained some information we
understood and some we didn't.
\begin{itemize}
\item
The most easily understood source of allocation
is the function \verb+vee''+, which is charged with
allocating about 66MB.
What \verb+vee''+ does is call memo combinators,
so the 66MB might be the memo table.
To~evaluate this hypothesis, we~estimated the amount of memory that
might be consumed by a memo-table entry.

@@

To~try to eliminate the unwanted allocations,
we~explored many variations on the code, of~which the most ambitious (and
most tedious) was to desugar the list comprehension by~hand and to
replace standard list operations with counterparts that participate
in stream fusion \cite{coutts:fusion}.
Nothing~helped.
\end{itemize}
In~desperation, we~tried to diagnose our problems by reading 
Core output from~GHC.
We~were guided by Don Stewart's invaluable blog posts,
and we also used his \texttt{ghc-core} tool, 
but it wasn't enough.
Even though one of us has worked on several optimizing compilers, 
we~just don't have enough experience to use GHC Core to diagnose
allocation problems in a 25-line function.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
to tests, QuickCheck properties, and generators.
Neither Matt nor SMURF includes test code.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
\mrfy's initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three tools were written
 in~haste.
We~have experience modifying all three tools.

\ifpagetuning\enlargethispage{\baselineskip}\fi

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was \emph{too} clever: 
nearly every change resulted in new~segfaults.

@@

We~built MRFy quickly, and we expected that
higher-order functions would make MRFy easy to extend.
Each new addition to MRFy's stochastic search took at most a day
to implement.
However, the next extension we wished to make
was based on
Richard Senington's model of \emph{local
search} \cite{senington:combinators-local}, 
which does not fit our abstraction of stochastic search.
We~wound up writing a new program that combines parts
of MRFy with Senington's  \texttt{local-search} package.
This~task took only three weeks, and the technical results were terrific.
Local search outperforms random hill-climbing, simulated
annealing, and the genetic algorithm.
Even though we're not entirely happy with the
state of the resulting code, we~count the story as a success.

Haskell encourages hasty programmers to slow down.
We~have to get the types right,
which makes it hard to write very large functions.
To~get the code to typecheck, we~have to write type signatures, which
also serve as documentation.
And once the types are accepted by the compiler,
it~is not much more work to write contracts for important functions.
MRFy is still hasty work.
Many types could be simplified;
we're~sure we've missed opportunities for abstraction;
and we know that MRFy's decomposition into modules could be improved.
But despite being hasty programmers, we produced code 
that is easy to understand and easy to extend.
Our hastily written Haskell beats
our hastily written Ruby and~C++.
